Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-05-14 survey/survey.tex
went over sec 1 and made small changes

What kind of operations can computers perform? This question was
answered several times in the last hundred years, where each answer
proposes an abstract \emph{model of computation} that specifies
allowable operations and (usually) their cost. The emerging consensus,
reflected in both early models of computations such as the Turing
Machine and the RAM as well as in the early Von Neumann and in modern
computer architectures, is that basic computer operations include
logical operations like conjunction, disjunction, and negation, as
well as reading from and writing to a large (infinite) collection of
memory locations.

No doubt, this consensus has been successful. And no doubt, these
operations \emph{can} indeed be performed on a computer. Yet, today,
with a possible quantum computing revolution in sight and with
unprecedented explosion in embedded computers and cyber-physical
systems, there are reasons to re-think this foundational question
again. In fact, the calls to re-think this foundational question have
been proclaimed by physicists more than forty years ago as the
following two quotes testify:

@@

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny `intelligence' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier: the formal notion of computation
(through the work of Turing and Church, 1936) and the formal notion of
information (through the work of Shannon, 1948). Landauer reasoned
that the computation done by the finite brain of the demon, involves
getting information about the movement of molecules, acting on that
information, and then overwriting it to make room for the next
computation.  In other words, the computation that is manipulating
information in the demon's brain \textit{must be thermodynamic work},
thereby bringing the demon back into the fold of physics.

This is a strange and wonderful idea: information, physics, and
computation are inextricably linked. In contrast, when the early
models of computation were developed, there was no compelling reason
to the take the information content of computations into consideration
-- in fact, at that time there was no quantifiable notion of
information. These models followed in the footsteps of logic where,
following hundreds of years of tradition, conventional operations such
conjunction and disjunction lose information by logically
irreversible. These models were already radical enough, by being
purely \emph{constructive}, in an era where classical mathematics,
with its pervasive use of excluded middle and frequent use of the
axiom of choice, did not overly worry about effective
representation. Landauer's observation implied however that ideas in
each field have consequences for the
other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
really appreciate this fact, we delve deeper into the origin of our
computational models and argue that they are essentially reflections
of contemporary laws of physics.

%%%%%%%%%
\paragraph*{Origins of Computational Models.}
Our current logical operations in high-level programming languages and
in our current hardware are both based on Boolean Logic -- which
really ought to be called \emph{Piercean logic}, as Boole had a rather
different conception of logic which is rather far from our current
models.  Nevertheless, going back to Boole's 1853 book entitled
\emph{An Investigation of the Laws of Thought, on which are Founded
  the Mathematical Theories of Logic and Probabilities}. The opening
sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
which clearly identifies the ``source'' of the logical laws as
mirroring Boole's understanding of human reasoning. A few chapters
later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
This ``law'' is reasonable in a classical world but is violated by the
postulates of quantum mechanics. Although a detailed historical
analysis of Boole's ideas in the light of modern physics is beyond our
scope, the above quotes should convey the idea that our elementary
computing notions date back to ideas that were thought reasonable
before \textasciitilde 1900.

@@

There are, in fact, many different quantum mechanical principles which
are at odds with our current models of computation. In this paper, we
will focus on the following particular aspect. We will view data as an
explicit representation of \emph{information} and programs as
processes that transform information in a reversible way, i.e.,
processes that are subject to the physical principle of
\emph{conservation of information.} We will formalize this idea and
follow its consequences, which will turn out to be far reaching.

@@

Starting from the physical perspective, whatever our notion of data
is, we will be interested in programs as representing transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of program. Of course, when programs themselves
are data, things do get more complicated.  In the following sections,
we will look at different natural classes of data, and explore the
corresponding notion of reversible programs.

@@

We now have in our hands our desired denotational semantics. We want
to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is completely
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language. Note how we have continued with
the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
but rather than being witnesses of inhabitation, they are witnesses of
equivences. Thus the proof language of commutative semiring equivalences
becomes terms of $\Pi$.

@@

In the previous section, we saw how we could take the proof
terms of commutative semirings as our starting point. What we need
now is to understand how proofs of algeraic identities should be
considered equivalent. Classical algebra does not help, as proofs
in algebra are not considered first-class citizens. However,
another route is available to us: since the work of 
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  Previously, we
modeled types as (finite) sets.

>>>>

What kind of operations can computers perform? This question was
answered several times in the last hundred years, where each answer
proposes an abstract \emph{model of computation} that specifies
allowable operations and (usually) their cost. The emerging consensus,
reflected in both early models of computations such as the Turing
Machine and the RAM as well as in the early Von Neumann and in modern
computer architectures, is that basic computer operations include
logical operations like conjunction, disjunction, and negation, as
well as reading from and writing to a large (infinite) collection of
memory locations. From this small set of primitive operations emerges
all higher-level programming languages and abstractions.

No doubt, this consensus on the available primitive physical
operations has been successful. And no doubt, these operations
\emph{can} indeed be performed on a computer. Yet, today, with a
possible quantum computing revolution in sight and with unprecedented
explosion in embedded computers and cyber-physical systems, there are
reasons to re-think this foundational question again. In fact, the
calls to re-think this foundational question have been proclaimed by
physicists more than forty years ago as the following two quotes
testify:

@@

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny ``intelligence'' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier: the formal notion of computation
(through the work of Turing and Church, 1936) and the formal notion of
information (through the work of Shannon, 1948). Landauer reasoned
that the computation done by the finite brain of the demon, involves
getting information about the movement of molecules, storing that
information, analyzing that information to act on it, and then --- and
this is the critical step --- overwriting it to make room for the next
computation.  In other words, the computation that is manipulating
information in the demon's brain \textit{must be thermodynamic work},
thereby bringing the demon back into the fold of physics.

This is a strange and wonderful idea: information, physics, and
computation are inextricably linked. In contrast, when the early
models of computation were developed, there was no compelling reason
to the take the information content of computations into consideration
-- in fact, at that time there was no quantifiable notion of
information. These models followed in the footsteps of logic where,
following hundreds of years of tradition, the truth of a statement was
seen as ``absolute'' and independent of any reasoning, understanding,
or action. Statements were either true or false with no regard to any
``observer'' and the idea that statements had information content that
should be preserved was outside the classical understanding of
logic. Hence the fact that conventional logic operations such as
conjunction and disjunction were logically irreversible and hence lose
information was not a concern. Landauer's observation implied however
that ideas in each field have consequences for the
other~\cite{bennett:1973:lrc,bennett1985fundamental,bennett2010notes,bennett2003notes,baker:1992:nft,baez2011physics,dblp:conf/csfw/malacarias12}. To
really appreciate this fact, we delve deeper into the origin of our
computational models and argue that they are essentially reflections
of contemporary laws of physics.

%%%%%%%%%
\paragraph*{Origins of Computational Models.}
Current high-level programming languages as well as current hardware
are both based on Boolean Logic -- which really ought to be called
\emph{Piercean logic}, as Boole had a rather different conception of
logic which is rather far from our current models.  Nevertheless,
going back to Boole's 1853 book entitled \emph{An Investigation of the
  Laws of Thought, on which are Founded the Mathematical Theories of
  Logic and Probabilities}, we find that the opening sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
which clearly identifies the ``source'' of the logical laws as
mirroring Boole's understanding of human reasoning. A few chapters
later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
This ``law'' is reasonable in a classical world but is violated by the
postulates of quantum mechanics. Although a detailed historical
analysis of Boole's ideas in the light of modern physics is beyond our
scope, the above quotes should convey the idea that our elementary
computing notions date back to ideas that were thought reasonable in
the late 1800s.

@@

There are, in fact, many different quantum mechanical principles which
are at odds with our current models of computation. In this paper, we
will focus on the previously identified principle of
\emph{reversibility}. In more detail, we will view data as an explicit
representation of \emph{information} and programs as processes that
transform information in a reversible way, i.e., processes that are
subject to the physical principle of \emph{conservation of
  information.} We will formalize this idea and follow its
consequences, which will turn out to be far reaching.

@@

Starting from the physical perspective, whatever our notion of data
is, we will be interested in programs as representing transformations
of that data which are reversible. In other words, we want our
programs-as-transformations to ``play well'' with the inherent notion
of ``sameness'' that our data will carry. Thus we need to start by
looking at what structure our data has, and that will help us define
an appropriate notion of a reversible program. Of course, when
programs themselves are data, things do get more complicated.  In the
following sections, we will look at different natural classes of data,
and explore the corresponding notion of reversible programs.

@@

We now have in our hands our desired denotational semantics. We want
to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is completely
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language. Note how we have continued with
the spirit of Curry-Howard: the terms of $\Pi$ are \emph{proof terms},
but rather than being witnesses of inhabitation, they are witnesses of
equivalences. Thus the proof language of commutative semiring equivalences
becomes terms of $\Pi$.

@@

In the previous section, we saw how we could take the proof
terms of commutative semirings as our starting point. What we need
now is to understand how proofs of algebraic identities should be
considered equivalent. Classical algebra does not help, as proofs
in algebra are not considered first-class citizens. However,
another route is available to us: since the work of 
Hofmann and Streicher~\cite{hofmann96thegroupoid}, we know that
one can model types as \emph{groupoids}.  Previously, we
modeled types as (finite) sets.
